---
title: "ACCC Airport Monitoring Report 2023-24"
subtitle: "Exploratory analysis of supplementary database — BNE, MEL, PER, SYD"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
    code-fold: true
    fig-width: 9
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)

accc_data <- readRDS("data_raw/accc_data.rds")

get_metric <- function(data, metric_pattern, cat = NULL) {
  out <- data |> filter(str_detect(metric, metric_pattern))
  if (!is.null(cat)) out <- out |> filter(category == cat)
  out
}
```

## Data overview

The dataset covers **4 monitored airports** (BNE, MEL, PER, SYD) across **20 financial years** (2004-05 to 2023-24), sourced from the ACCC Airport Monitoring Report 2023-24 supplementary database.

```{r}
#| label: overview-table

accc_data |>
  summarise(
    Rows       = scales::comma(n()),
    Airports   = n_distinct(airport),
    Categories = n_distinct(category),
    Metrics    = n_distinct(metric),
    Years      = n_distinct(year),
    `NA (%)` = round(mean(is.na(value)) * 100, 1)
  ) |>
  knitr::kable()
```

```{r}
#| label: overview-counts

accc_data |>
  count(airport, category) |>
  pivot_wider(names_from = category, values_from = n) |>
  rename(Airport = airport,
         `Total/Aero rows` = total_aero,
         `Carpark/Landside rows` = carpark_landside) |>
  knitr::kable()
```

---

## Passenger volumes

### Total passengers over time

```{r}
#| label: fig-pax-total
#| fig-cap: "Total passengers by airport, 2004-05 to 2023-24"

passengers <- get_metric(accc_data, "^Passengers \\('000\\)$", "total_aero")

ggplot(passengers, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  geom_point(size = 1.5) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = NULL, y = "Passengers ('000)", colour = "Airport")
```

### Domestic vs International split (2023-24)

```{r}
#| label: fig-pax-split
#| fig-cap: "Domestic vs International passengers by airport, 2023-24"

dom_intl <- accc_data |>
  filter(
    metric %in% c("Domestic Passengers ('000)", "International Passengers ('000)"),
    category == "total_aero",
    year == "2023-24"
  )

ggplot(dom_intl, aes(x = airport, y = value, fill = metric)) +
  geom_col(position = "dodge") +
  labs(x = "Airport", y = "Passengers ('000)", fill = NULL)
```

### COVID-19 impact summary

```{r}
#| label: tbl-covid
#| tbl-cap: "Passenger volumes and COVID-19 impact by airport"

passengers |>
  filter(year %in% c("2018-19", "2019-20", "2020-21", "2021-22", "2022-23", "2023-24")) |>
  select(airport, year, value) |>
  pivot_wider(names_from = year, values_from = value) |>
  mutate(
    `Drop to 20-21 (%)` = round((`2020-21` - `2018-19`) / `2018-19` * 100, 1),
    `Recovery by 22-23 (%)` = round((`2022-23` - `2018-19`) / `2018-19` * 100, 1),
    `Recovery by 23-24 (%)` = round((`2023-24` - `2018-19`) / `2018-19` * 100, 1)
  ) |>
  rename(Airport = airport) |>
  knitr::kable()
```

---

## Aeronautical financials

### Revenue, expenses and profit

```{r}
#| label: fig-aero-fin
#| fig-cap: "Aeronautical revenue, expenses and profit by airport"
#| fig-height: 9

aero_financials <- accc_data |>
  filter(
    category == "total_aero",
    str_detect(
      metric,
      regex(
        "^aeronautical (revenue|operating expenses|operating profit) \\(\\$m\\)$",
        ignore_case = TRUE
      )
    )
  )

ggplot(aero_financials, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "$million", colour = "Airport")
```

### Profit margin

```{r}
#| label: fig-aero-margin
#| fig-cap: "Aeronautical profit margin by airport (%)"

aero_margin <- accc_data |>
  filter(
    category == "total_aero",
    str_detect(metric, regex("^aeronautical profit margin \\(%\\)$", ignore_case = TRUE))
  )

ggplot(aero_margin, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = NULL, y = "Profit margin (%)", colour = "Airport")
```

### EBITA rate of return — aeronautical services

```{r}
#| label: fig-ebita-aero
#| fig-cap: "Rate of return (EBITA) on tangible non-current assets — aeronautical services (%)"

ebita_aero <- accc_data |>
  filter(
    category == "total_aero",
    metric == "Rate of return (EBITA) on tangible non-current assets for aeronautical services (%)"
  )

ggplot(ebita_aero, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = NULL, y = "Rate of return (%)", colour = "Airport")
```

### EBITA rate of return — total airport services

```{r}
#| label: fig-ebita-total
#| fig-cap: "Rate of return (EBITA) on tangible non-current assets — total airport services, excluding inter-company (%)"

ebita_total <- accc_data |>
  filter(
    category == "total_aero",
    metric == "Rate of return (EBITA) on tangible non-current assets for total airport services (%) - excluding inter-company"
  )

ggplot(ebita_total, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = NULL, y = "Rate of return (%)", colour = "Airport")
```

### Per-passenger metrics (2023-24)

```{r}
#| label: fig-per-pax
#| fig-cap: "Aeronautical per-passenger metrics by airport, 2023-24"

per_pax <- accc_data |>
  filter(
    category == "total_aero",
    metric %in% c(
      "Aeronautical Revenue per passenger ($)",
      "Aeronautical Expenses per passenger ($)",
      "Aeronautical Operating profit per passenger ($)"
    ),
    year == "2023-24"
  )

ggplot(per_pax, aes(x = airport, y = value, fill = airport)) +
  geom_col() +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = NULL, y = "$", fill = "Airport")
```

---

## Total airport financials

### Revenue, expenses and profit

```{r}
#| label: fig-total-fin
#| fig-cap: "Total airport revenue, expenses and profit by airport"
#| fig-height: 9

total_financials <- accc_data |>
  filter(
    category == "total_aero",
    metric %in% c(
      "Total airport revenue ($m)",
      "Total airport operating expenses ($m)",
      "Total airport operating profit ($m)"
    )
  )

ggplot(total_financials, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "$million", colour = "Airport")
```

### Rate of return on aeronautical assets

```{r}
#| label: fig-ror
#| fig-cap: "Rate of return (EBITA) on aeronautical tangible assets (%)"

ror <- get_metric(
  accc_data,
  "^Rate of return \\(EBITA\\) on tangible non-current assets for aeronautical services \\(%\\)$",
  "total_aero"
)

ggplot(ror, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = NULL, y = "Rate of return (%)", colour = "Airport")
```

---

## Car park financials

### Revenue, expenses and profit

```{r}
#| label: fig-carpark-fin
#| fig-cap: "Car parking revenue, expenses and profit by airport"
#| fig-height: 9

carpark_fin <- accc_data |>
  filter(
    category == "carpark_landside",
    metric %in% c(
      "Car parking revenue ($million)",
      "Car parking operating expenses ($million)",
      "Car parking operating profit ($million)"
    )
  )

ggplot(carpark_fin, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "$million", colour = "Airport")
```

### Profit margin

```{r}
#| label: fig-carpark-margin
#| fig-cap: "Car parking profit margin by airport (%)"

cp_margin <- get_metric(accc_data, "^Car parking profit margin", "carpark_landside")

ggplot(cp_margin, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = NULL, y = "Profit margin (%)", colour = "Airport")
```

### Revenue per vehicle

```{r}
#| label: fig-rev-per-veh
#| fig-cap: "Car parking revenue per vehicle by airport ($)"

rev_per_veh <- get_metric(accc_data, "^Revenue per vehicle", "carpark_landside")

ggplot(rev_per_veh, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = NULL, y = "$", colour = "Airport")
```

---

## Landside access revenue

```{r}
#| label: fig-landside
#| fig-cap: "Total landside access revenue by airport ($ millions)"

landside_rev <- get_metric(accc_data, "^Total landside access revenue", "carpark_landside")

ggplot(landside_rev, aes(x = year, y = value, colour = airport, group = airport)) +
  geom_line() +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = NULL, y = "$ millions", colour = "Airport")
```
