---
title: "Australian Domestic Aviation — On-Time Performance"
subtitle: "BITRE OTP Time Series, January 2010 – January 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    embed-resources: true
    code-fold: true
    theme: flatly
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| code-fold: false

library(tidyverse)
library(lubridate)
library(scales)
library(knitr)

otp_data    <- readRDS("data_raw/otp_data.rds")
airport_mvt <- readRDS("data_raw/airport_movements.rds")

# Major scheduled carriers — excludes the "All Airlines" aggregate row
major_airlines <- c(
  "Qantas",
  "Jetstar",
  "Virgin Australia",
  "Virgin Australia Regional Airlines",
  "Tigerair Australia",
  "Rex Airlines",
  "Regional Express"
)

otp_major <- otp_data |>
  filter(airline %in% major_airlines)

# Airport-level aggregate (all airlines combined) for airport comparisons
accc_ports <- c("Sydney", "Melbourne", "Brisbane", "Perth")

otp_by_airport <- otp_data |>
  filter(airline == "All Airlines", departing_port %in% accc_ports) |>
  summarise(
    sectors_scheduled  = sum(sectors_scheduled,  na.rm = TRUE),
    sectors_flown      = sum(sectors_flown,       na.rm = TRUE),
    cancellations      = sum(cancellations,       na.rm = TRUE),
    departures_on_time = sum(departures_on_time,  na.rm = TRUE),
    .by = c(departing_port, date, year, month)
  ) |>
  mutate(
    pct_on_time  = departures_on_time / sectors_flown      * 100,
    pct_cancelled = cancellations     / sectors_scheduled  * 100
  )
```

## Data overview

The BITRE On-Time Performance series covers scheduled domestic jet services
operated by Australia's major carriers. It records sectors scheduled and flown,
cancellations, and on-time arrivals and departures (within 15 minutes of
schedule) at the route × airline × month level.

```{r}
#| label: tbl-coverage
#| tbl-cap: "Date coverage by airline"

otp_data |>
  filter(airline %in% major_airlines) |>
  summarise(
    From  = min(date),
    To    = max(date),
    Years = n_distinct(year),
    .by   = airline
  ) |>
  arrange(From) |>
  rename(Airline = airline) |>
  kable()
```

---

## Airline capacity — sectors flown

### Annual sectors flown by airline

Qantas Group (Qantas + Jetstar) dominates domestic capacity, with Virgin
Australia as the principal competitor. Tigerair Australia's entry in 2007 and
eventual absorption into Virgin in 2020 is visible as a brief capacity addition.
Virgin Australia's collapse into voluntary administration in April 2020, and its
subsequent restart as a smaller carrier, is the defining structural event in the
series.

```{r}
#| label: fig-sectors-airline
#| fig-cap: "Annual sectors flown by airline"

otp_major |>
  filter(route != "All Ports-All Ports") |>
  summarise(
    sectors_flown = sum(sectors_flown, na.rm = TRUE),
    .by = c(airline, year)
  ) |>
  ggplot(aes(x = year, y = sectors_flown, colour = airline)) +
  geom_line() +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Sectors flown", colour = "Airline")
```

### Domestic market share — sectors flown

```{r}
#| label: fig-market-share
#| fig-cap: "Domestic market share by sectors flown (annual)"

otp_major |>
  summarise(
    sectors_flown = sum(sectors_flown, na.rm = TRUE),
    .by = c(airline, year)
  ) |>
  mutate(share = sectors_flown / sum(sectors_flown), .by = year) |>
  ggplot(aes(x = year, y = share, fill = airline)) +
  geom_area() +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "Share of sectors", fill = "Airline")
```

---

## On-time performance

### On-time departure performance by airline

The dashed line marks the 80% threshold that is commonly used as a benchmark
for acceptable punctuality. Performance dipped industry-wide in 2022–23 as
airlines ramped capacity rapidly while facing crew shortages and airport
congestion in the post-COVID restart.

```{r}
#| label: fig-otp-airline
#| fig-cap: "On-time departure performance by airline (annual)"

otp_major |>
  summarise(
    sectors_flown      = sum(sectors_flown,       na.rm = TRUE),
    departures_on_time = sum(departures_on_time,  na.rm = TRUE),
    .by = c(airline, year)
  ) |>
  mutate(pct_on_time = departures_on_time / sectors_flown * 100) |>
  ggplot(aes(x = year, y = pct_on_time, colour = airline)) +
  geom_line() +
  geom_hline(yintercept = 80, linetype = "dashed", colour = "grey50") +
  labs(
    subtitle = "Dashed line = 80% benchmark",
    x = NULL, y = "On-time departures (%)", colour = "Airline"
  )
```

### Cancellation rate by airline

```{r}
#| label: fig-cancellations
#| fig-cap: "Cancellation rate by airline (annual)"

otp_major |>
  summarise(
    sectors_scheduled = sum(sectors_scheduled, na.rm = TRUE),
    cancellations     = sum(cancellations,     na.rm = TRUE),
    .by = c(airline, year)
  ) |>
  mutate(pct_cancelled = cancellations / sectors_scheduled * 100) |>
  ggplot(aes(x = year, y = pct_cancelled, colour = airline)) +
  geom_line() +
  labs(x = NULL, y = "Cancellations (%)", colour = "Airline")
```

---

## Airport-level performance

### On-time departures by airport (all airlines)

Airport-level OTP reflects a mix of airline punctuality and airport
infrastructure constraints. Sydney consistently underperforms relative to other
capitals, partly due to the single runway constraint during peak periods.

```{r}
#| label: fig-otp-airport
#| fig-cap: "On-time departure performance — ACCC airports, all airlines (monthly with smoothed trend)"

otp_by_airport |>
  ggplot(aes(x = date, y = pct_on_time, colour = departing_port)) +
  geom_line(alpha = 0.3) +
  geom_smooth(se = FALSE, span = 0.15) +
  labs(
    subtitle = "Raw monthly (faint) + smoothed trend",
    x = NULL, y = "On-time departures (%)", colour = "Airport"
  )
```

---

## Bridging OTP and BITRE movements data

### OTP sectors flown vs BITRE domestic outbound movements

The OTP series records scheduled domestic departures only. BITRE
`dom_outbound` is the correct comparator (both are departure counts), but
includes freight, charter and general aviation not captured in OTP. The
residual gap of roughly 10–20% varies by airport — Perth's larger gap reflects
its significant resource-sector charter traffic.

```{r}
#| label: fig-otp-vs-bitre
#| fig-cap: "OTP scheduled sectors vs BITRE domestic outbound movements — ACCC airports (monthly)"
#| fig-height: 7

port_to_airport <- c(
  "Sydney"    = "SYDNEY",
  "Melbourne" = "MELBOURNE",
  "Brisbane"  = "BRISBANE",
  "Perth"     = "PERTH"
)

otp_sectors_monthly <- otp_data |>
  filter(airline == "All Airlines") |>
  summarise(
    sectors_scheduled = sum(sectors_scheduled, na.rm = TRUE),
    sectors_flown     = sum(sectors_flown,     na.rm = TRUE),
    .by = c(departing_port, date)
  )

otp_vs_mvt <- otp_sectors_monthly |>
  filter(departing_port %in% names(port_to_airport)) |>
  mutate(airport = port_to_airport[departing_port]) |>
  left_join(
    airport_mvt |> select(airport, date, dom_outbound),
    join_by(airport, date)
  )

otp_vs_mvt |>
  pivot_longer(
    c(sectors_flown, dom_outbound),
    names_to = "source", values_to = "movements"
  ) |>
  mutate(
    source = if_else(
      source == "sectors_flown",
      "OTP (scheduled domestic departures)",
      "BITRE (all domestic departures)"
    )
  ) |>
  ggplot(aes(x = date, y = movements, colour = source)) +
  geom_line(alpha = 0.7) +
  scale_y_continuous(labels = comma) +
  facet_wrap(~departing_port, scales = "free_y") +
  labs(
    subtitle = "Gap ≈ freight, charter & general aviation not in OTP",
    x = NULL, y = "Movements / sectors", colour = "Source"
  )
```

---

## Virgin Australia — case study

### OTP before, during and after administration

Virgin Australia entered voluntary administration on 21 April 2020 and was
acquired by Bain Capital in November 2020. The airline restarted as a leaner
operation with a smaller fleet and network. On-time performance improved
markedly in 2021 as the restart benefited from reduced congestion and a
simplified network before deteriorating again with the industry-wide post-COVID
operational squeeze.

```{r}
#| label: fig-va-otp
#| fig-cap: "Virgin Australia — on-time departure performance (all VA brands combined)"

va_otp <- otp_data |>
  filter(str_detect(airline, "Virgin Australia")) |>
  summarise(
    sectors_flown      = sum(sectors_flown,       na.rm = TRUE),
    departures_on_time = sum(departures_on_time,  na.rm = TRUE),
    cancellations      = sum(cancellations,       na.rm = TRUE),
    sectors_scheduled  = sum(sectors_scheduled,   na.rm = TRUE),
    .by = date
  ) |>
  mutate(
    pct_on_time   = departures_on_time / sectors_flown     * 100,
    pct_cancelled = cancellations      / sectors_scheduled * 100,
    era = case_when(
      date < as.Date("2020-04-01") ~ "Pre-administration",
      date < as.Date("2021-01-01") ~ "Administration / restart",
      TRUE                         ~ "Post-administration"
    )
  )

ggplot(va_otp, aes(x = date, y = pct_on_time, colour = era, group = 1)) +
  geom_line() +
  geom_vline(xintercept = as.Date("2020-04-01"),
             linetype = "dashed", colour = "grey40") +
  scale_colour_manual(
    values = c(
      "Pre-administration"      = "#2166ac",
      "Administration / restart" = "#d7191c",
      "Post-administration"     = "#1a9641"
    )
  ) +
  labs(
    subtitle = "Dashed line = voluntary administration (Apr 2020)",
    x = NULL, y = "On-time departures (%)", colour = NULL
  )
```

---

## Top routes

### Most-flown domestic routes

```{r}
#| label: fig-top-routes
#| fig-cap: !expr paste0("Top 15 routes by sectors flown — ", max(otp_data$year) - 1)

otp_data |>
  filter(
    airline == "All Airlines",
    year    == max(year) - 1,
    route   != "All Ports-All Ports"
  ) |>
  summarise(sectors_flown = sum(sectors_flown, na.rm = TRUE), .by = route) |>
  slice_max(sectors_flown, n = 15) |>
  mutate(route = fct_reorder(route, sectors_flown)) |>
  ggplot(aes(x = sectors_flown, y = route)) +
  geom_col(fill = "#2166ac") +
  scale_x_continuous(labels = comma) +
  labs(x = "Sectors flown", y = NULL)
```
