---
title: "Australian Airport Traffic — Monthly Analysis"
subtitle: "BITRE WebMonthly Airport Statistics, January 2009 – September 2025"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    embed-resources: true
    code-fold: true
    theme: flatly
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| code-fold: false

library(tidyverse)
library(lubridate)
library(scales)
library(knitr)

airport_pax <- readRDS("data_raw/airport_pax.rds")
airport_mvt <- readRDS("data_raw/airport_movements.rds")
otp_data    <- readRDS("data_raw/otp_data.rds")

# Consistent ordering for the four ACCC-monitored airports
airport_order <- c("SYDNEY", "MELBOURNE", "BRISBANE", "PERTH")

accc_airports <- airport_pax |>
  filter(airport %in% airport_order) |>
  mutate(airport = factor(airport, levels = airport_order))

accc_mvt <- airport_mvt |>
  filter(airport %in% airport_order) |>
  mutate(airport = factor(airport, levels = airport_order))
```

## Data overview

The BITRE WebMonthly series covers **`r n_distinct(airport_pax$airport)`
airports** from **`r format(min(airport_pax$date), "%B %Y")`** to
**`r format(max(airport_pax$date), "%B %Y")`**, recording inbound, outbound and
total passenger movements together with aircraft movement counts for domestic
and international services. The table below summarises the most recent complete
calendar year.

```{r}
#| label: tbl-overview
#| tbl-cap: "Airport traffic summary — calendar year 2024"

last_year <- year(max(airport_pax$date)) - 1  # last complete year

airport_pax |>
  filter(airport != "All Australian Airports", year == last_year) |>
  summarise(
    total_pax  = sum(total_total, na.rm = TRUE),
    dom_pax    = sum(dom_total,   na.rm = TRUE),
    intl_pax   = sum(intl_total,  na.rm = TRUE),
    .by = airport
  ) |>
  left_join(
    airport_mvt |>
      filter(airport != "All Australian Airports", year == last_year) |>
      summarise(
        dom_mvt  = sum(dom_total,  na.rm = TRUE),
        intl_mvt = sum(intl_total, na.rm = TRUE),
        .by = airport
      ),
    join_by(airport)
  ) |>
  arrange(desc(total_pax)) |>
  mutate(
    intl_share = scales::percent(intl_pax / total_pax, accuracy = 0.1),
    across(total_pax:intl_pax, ~ scales::comma(round(.x / 1e6, 2), suffix = "m")),
    across(dom_mvt:intl_mvt,   ~ scales::comma(.x))
  ) |>
  rename(
    Airport = airport,
    `Total pax` = total_pax, `Dom pax` = dom_pax, `Intl pax` = intl_pax,
    `Intl share` = intl_share,
    `Dom movements` = dom_mvt, `Intl movements` = intl_mvt
  ) |>
  kable()
```

---

## Passenger volumes

### Total monthly passengers — ACCC airports

Sydney consistently leads on total passengers, with Melbourne close behind.
The sharp collapse in April 2020 from COVID-19 border closures and the
subsequent recovery pattern are clearly visible across all four airports.

```{r}
#| label: fig-total-pax
#| fig-cap: "Total monthly passengers — ACCC airports (Jan 2009 – Sep 2025)"

accc_airports |>
  ggplot(aes(x = date, y = total_total, colour = airport)) +
  geom_line(alpha = 0.6) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Passengers", colour = "Airport")
```

### Domestic vs international split

International traffic is proportionally most important at Sydney and Perth.
Perth's international share reflects significant Asia-Pacific routes and
fly-in fly-out resource sector traffic that sits outside the scheduled
domestic market.

```{r}
#| label: fig-dom-intl
#| fig-cap: "Domestic vs international passengers — ACCC airports"
#| fig-height: 7

accc_airports |>
  select(airport, date, dom_total, intl_total) |>
  pivot_longer(c(dom_total, intl_total), names_to = "type", values_to = "pax") |>
  mutate(type = if_else(type == "dom_total", "Domestic", "International")) |>
  ggplot(aes(x = date, y = pax, colour = type)) +
  geom_line(alpha = 0.7) +
  scale_y_continuous(labels = comma) +
  facet_wrap(~airport, scales = "free_y") +
  labs(x = NULL, y = "Passengers", colour = NULL)
```

### International share over time

Sydney's international share is structurally higher (~35%) than Brisbane and
Melbourne (~23–25%). Perth's share rose through the 2010s as Asian routes
expanded, and has recovered post-COVID more slowly than domestic.

```{r}
#| label: fig-intl-share
#| fig-cap: "International share of total passengers — ACCC airports"

accc_airports |>
  mutate(intl_share = intl_total / total_total) |>
  ggplot(aes(x = date, y = intl_share, colour = airport)) +
  geom_line(alpha = 0.6) +
  scale_y_continuous(labels = percent) +
  labs(x = NULL, y = "International share", colour = "Airport")
```

### Annual volumes — all reporting airports

```{r}
#| label: fig-all-airports
#| fig-cap: "Annual total passengers — all reporting airports"
#| fig-height: 7

airport_pax |>
  filter(airport != "All Australian Airports") |>
  summarise(total_pax = sum(total_total, na.rm = TRUE), .by = c(airport, year)) |>
  mutate(airport = fct_reorder(airport, total_pax, .fun = sum, .desc = TRUE)) |>
  ggplot(aes(x = year, y = total_pax, colour = airport)) +
  geom_line() +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Passengers", colour = "Airport")
```

---

## Aircraft movements

### Domestic vs international outbound movements

Aircraft movement counts follow passenger trends closely, but the ratio of
passengers per movement (see next section) reveals changes in fleet mix and
load factors over time.

```{r}
#| label: fig-movements
#| fig-cap: "Domestic vs international outbound movements — ACCC airports"
#| fig-height: 7

accc_mvt |>
  select(airport, date, dom_outbound, intl_outbound) |>
  pivot_longer(c(dom_outbound, intl_outbound),
               names_to = "type", values_to = "movements") |>
  mutate(type = if_else(type == "dom_outbound", "Domestic", "International")) |>
  ggplot(aes(x = date, y = movements, colour = type)) +
  geom_line(alpha = 0.7) +
  scale_y_continuous(labels = comma) +
  facet_wrap(~airport, scales = "free_y") +
  labs(x = NULL, y = "Aircraft movements", colour = NULL)
```

### Passengers per departure (load proxy)

Dividing passengers by outbound movements gives an implied passengers-per-
departure figure. This is not a true load factor (no seat count is available)
but is a useful proxy for the combined effect of larger aircraft and higher
load factors. The upward trend pre-COVID reflects fleet upgrades (e.g. 737
replacing Fokker, A320 family growth) and improved yield management.

```{r}
#| label: fig-pax-per-dep
#| fig-cap: "Passengers per departure (load proxy) — ACCC airports"
#| fig-height: 5

accc_airports |>
  left_join(
    accc_mvt |> select(airport, date, mvt_dom_out = dom_outbound, mvt_intl_out = intl_outbound),
    join_by(airport, date)
  ) |>
  mutate(
    pax_per_dom_dep  = dom_total  / mvt_dom_out,
    pax_per_intl_dep = intl_total / mvt_intl_out
  ) |>
  pivot_longer(
    c(pax_per_dom_dep, pax_per_intl_dep),
    names_to = "type", values_to = "pax_per_dep"
  ) |>
  mutate(type = if_else(type == "pax_per_dom_dep", "Domestic", "International")) |>
  ggplot(aes(x = date, y = pax_per_dep, colour = airport)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~type, scales = "free_y") +
  labs(
    subtitle = "Higher = fuller planes and/or larger aircraft",
    x = NULL, y = "Passengers per departure", colour = "Airport"
  )
```

---

## Seasonality

### Pre-COVID domestic seasonality (2009–2019)

Domestic travel peaks in January (summer school holidays) and July (winter
school holidays), with a secondary peak in September–October. Brisbane and
Sydney show a more pronounced January peak than Melbourne, reflecting
stronger leisure and interstate holiday travel to Queensland.

```{r}
#| label: fig-seasonality
#| fig-cap: "Average domestic passengers by calendar month — pre-COVID (2009–2019)"

accc_airports |>
  filter(year < 2020) |>
  summarise(avg_pax = mean(dom_total, na.rm = TRUE), .by = c(airport, month)) |>
  mutate(month_label = month(month, label = TRUE)) |>
  ggplot(aes(x = month_label, y = avg_pax, colour = airport, group = airport)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Average passengers", colour = "Airport")
```

---

## COVID-19 impact and recovery

Indexing to January 2020 = 100 allows direct comparison of the depth of the
COVID collapse and the pace of recovery across airports. Perth recovered
domestic traffic fastest, aided by the reopening of intrastate WA routes.
Sydney and Melbourne, with higher international exposure, took longer to reach
pre-COVID levels.

```{r}
#| label: fig-covid
#| fig-cap: "Passenger recovery indexed to January 2020 = 100"

base_jan2020 <- accc_airports |>
  filter(year == 2020, month == 1) |>
  select(airport, base = total_total)

accc_airports |>
  filter(date >= as.Date("2019-01-01")) |>
  left_join(base_jan2020, join_by(airport)) |>
  mutate(index = total_total / base * 100) |>
  ggplot(aes(x = date, y = index, colour = airport)) +
  geom_line() +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey50") +
  labs(x = NULL, y = "Index (Jan 2020 = 100)", colour = "Airport")
```

---

## Bridging BITRE and OTP data

### Domestic passengers per OTP sector — ACCC airports

The OTP dataset records scheduled domestic departures by airline. Dividing
BITRE domestic passengers by OTP sectors flown gives a passengers-per-sector
series that blends load factor and aircraft size effects. Values above ~150
reflect the progressive transition to larger narrowbody aircraft (A320/737-800)
from the mid-2010s onward.

Note: OTP covers scheduled domestic services only; BITRE `dom_outbound` also
includes freight, charter and general aviation. The ratio therefore slightly
understates true passengers per scheduled departure.

```{r}
#| label: fig-pax-per-sector
#| fig-cap: "Domestic passengers per OTP sector — ACCC airports (annual)"

port_to_airport <- c(
  "Sydney"    = "SYDNEY",
  "Melbourne" = "MELBOURNE",
  "Brisbane"  = "BRISBANE",
  "Perth"     = "PERTH"
)

otp_annual <- otp_data |>
  filter(
    airline == "All Airlines",
    departing_port %in% names(port_to_airport)
  ) |>
  summarise(
    sectors_flown = sum(sectors_flown, na.rm = TRUE),
    .by = c(departing_port, year)
  ) |>
  mutate(airport = port_to_airport[departing_port])

# Sum all routes departing from each ACCC port (All Airlines aggregate)
otp_annual <- otp_data |>
  filter(
    airline == "All Airlines",
    departing_port %in% names(port_to_airport)
  ) |>
  summarise(
    sectors_flown = sum(sectors_flown, na.rm = TRUE),
    .by = c(departing_port, year)
  ) |>
  mutate(airport = port_to_airport[departing_port])

bitre_annual <- accc_airports |>
  summarise(dom_pax = sum(dom_total, na.rm = TRUE), .by = c(airport, year)) |>
  mutate(airport = as.character(airport))

bridge <- otp_annual |>
  left_join(bitre_annual, join_by(airport, year)) |>
  mutate(pax_per_sector = dom_pax / sectors_flown)

bridge |>
  ggplot(aes(x = year, y = pax_per_sector, colour = airport)) +
  geom_line() +
  geom_point() +
  labs(
    subtitle = "BITRE dom pax ÷ OTP sectors flown. Reflects aircraft size & load trends.",
    x = NULL, y = "Passengers per sector", colour = "Airport"
  )
```
